Abstractions for Type-Level Programming

### 2022-04-22 preprint.tex
Address Dragana's comments

\begin{diff}
A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
Although the ideas presented in this did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala 3 language.
\end{diff}

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
\begin{diff}
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
\footnotetext{%
When a generalized singleton type contains a component of form \texttt{(\_: T)}, then that type is not singleton: it might contains more than one value.}
\end{diff}

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type.
However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
\begin{diff}
More generally, when a dependently-typed function call into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
\end{diff}

@@

\begin{diff}
Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
His type system is much more expressive than ours and allows bounded quantification over both types and terms using the notion of a |Type| type and power types.
Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.

In~\citep{aspinall1994subtyping} Aspinall introduces a dependently-typed system with subtyping \emph{and} singleton types that resembles ours in its type language.
His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
Aspinall's system also has primitives types and allows for atomic subtyping among them, but no congruence rules, so partially-widened forms like |(_: Any) :: Nil| cannot be represented.

Pure subtype systems~\citep{hutchins2010pure} are different in that they only contain a single syntactic category for both terms and types, and a single relation, subtyping, that subsumes typing, subtyping and type evaluation of our system.
However, their system allows for partially-widened types similar to ours and also enables the computation with different levels of precision.
For instance, it is able to conclude that |Int + 5| can be approximated as |Int|.

Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
Though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their goals largely orthogonal to ours.
\end{diff}

>>>>

\begin{diff}
A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
Although the ideas presented in this chapter did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala language.
\end{diff}

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
\begin{diff}
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
\footnotetext{%
When a generalized singleton type contains a component of form \texttt{(\_: T)}, that type is not singleton: it might contains more than one value.}
\end{diff}

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type.
However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
\begin{diff}
More generally, when a dependently-typed function calls into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
\end{diff}

@@

\begin{diff}
Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
\citeauthor{cardelli1988structural}'s type system is much more expressive than ours and allows bounded quantification over both types and terms, using the notion of a |Type| type and power types.
Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.

In~\citep{aspinall1994subtyping} Aspinall introduces a dependently-typed system with subtyping \emph{and} singleton types which resembles ours in its type language.
His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
Aspinall's system also has primitive types and allows for atomic subtyping among them, but no congruence rules, so partially-widened forms like |(_: Any) :: Nil| cannot be represented.

Pure subtype systems~\citep{hutchins2010pure} are different in that they only contain a single syntactic category for both terms and types, and a single relation, subtyping, that subsumes typing, subtyping and type evaluation of our system.
However, \citeauthor{hutchins2010pure}'s system allows for partially-widened types similar to ours and also enables computations with different levels of precision.
For instance, it is able to conclude that |Int + 5| can be approximated as |Int|.

Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
Though they introduce a form of dependency, path-dependent types were not designed for type-level computations, rendering their goals largely orthogonal to ours.
\end{diff}
